flowCore a Bioconductor package for high throughput flow cytometry

### 2008-12-16 paper/bmc_article.tex
more changes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35947 bc3139a8-67e5-0310-9ffc-ced21a209358

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
  \paragraph*{Background:} Recent advances in automation technologies
  have enabled the use of flow-cytometry high content screening
  (FH-HCS). However, data management and data analysis methods have
  not advanced sufficiently far from the initial small-scale studies to
  support the large complex data sets generated in FH-HCS
  applications.

  \paragraph*{Results:} We developed a set of flexible computational
  tools for FC-HCS data analysis written in R and freely available
  from the Bioconductor Project. The \Rpackage{flowCore} package
  provides the flow cytometry community with a shared research
  platform that enables bioinformaticians, computer scientists, and
  statisticians to work collaboratively with biologists and clinicians
  to develop novel methods for flow cytometry data analysis.

  \paragraph*{Conclusions:} We hope that our framework will be the
  foundation for fruitful collaborative research involving multiple
  scientific fields and that it will help facilitate the development
  and deployment of FC-HCS methods to complex scientific and clinical
  applications.
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Automation technologies developed during the last several years have
enabled the use of flow cytometry high content screening (FC-HCS) to
generate large, complex datasets in both basic and clinical research
applications \cite{brinkman2007hcf}. A serious bottleneck in the
interpretation of existing studies and the application of FC-HCS to
even larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of flow cytometry (FCM) to small-scale,
tube-based studies \cite{mahnke2007omi}.
%% add one of the last roederer's paper that "request" algo and
%% automation for FC-HCS data analysis.

Some of the consequences of this lag are difficulties in maintaining
the integrity and documentation of extremely large datasets, assessing
measurement quality, developing validated assays, controlling the
accuracy of gating techniques, automating complex gating strategies,
and aggregating statistical results across large study sets for
further analysis. The most serious consequence, however, is that new
analysis approaches face difficulty in finding their way into standard
practice. We believe that this barrier to the development and
dissemination of new analysis methods is one of the fundamental
restraints on the future expansion of FC-HCS in both clinical and
research applications.

Another critical component is the development and implementation of
standards that will facilitate the adoption of these methods by both
the larger research community and commercial software vendors.
Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. Most of the currently available
analysis tools are designed to reflect this workflow.  We believe that
these approaches, in addition to being expensive and labor intensive,
do not fully address the highly complex nature of FCM data; in
particular, they disregard many of the fundamental aspects of the
data, such as its underlying distribution and its high-dimensional
nature. Furthermore, the subjective character of manual analyses are a
major obstacle to reproducibility. In a recent study of flow
cytometric standardization involving 15 institutions, the mean
inter-laboratory coefficient of variation ranged from 17 to 44\%, even
though preparation was standardized and performed using the same
samples and reagents at each site \cite{Maecker2005}. For FC-HCS data,
unassisted manual inspection is extremely time consuming, and robust
statistical methods need to be developed to point investigators to
interesting aspects of the data, or to potential problems. While the
expert knowledge of immunologists and researchers remains crucial for
the understanding of FCM data, we believe that collaboration with
other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

In this paper, we describe a set of flexible and well structured
computational tools to efficiently analyze FC-HCS data. This set of
tools is based on the open source statistical software R \cite{Rmain}
and distributed as the Bioconductor package \Rpackage{flowCore}
\cite{BIOC}. Our intent is to provide a shared research platform that
enables bioinformaticians, computer scientists, and statisticians to
work collaboratively with biologists and clinicians to develop novel
methods for FCM data analysis, a process deemed crucial by many for
the further development of the technology \cite{lizard2007fca}.

@@

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
The computational tools we have developed are distributed in the R
software language as the Bioconductor package \Rpackage{flowCore}.
The Bioconductor package \Rpackage{flowCore} is a freely available,
highly functional, and extensible FCM data analysis platform that
enables researchers to efficiently handle FC-HCS data and encourages
open development of tools for their coherent analysis. In our
implementation of \Rpackage{flowCore} we rely on two important lessons
learned from the field of gene expression data analysis: the first
being the importance of data structures that reflect the underlying
data and facilitate the manipulations that are of most interest, while
the second is the importance of a modular architecture that allows for
many developers to extend and use the underlying infrastructure and to
combine tools in complex work flows.  \Rpackage{flowCore} implements
such computationally efficient data structures and a range of
specialized methods addressing all components of a typical FCM
analysis work flow, including compensation, transformation, and
gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
operating systems. Conceptually, the underlying data structures that
build the core of our software can be implemented in any programming
language. We chose the R language for its advantages in statistical
programming and visualization.
%% In this paper, we describe the functionality and methods underlying
%% \Rpackage{flowCore}, and provide examples of their use.

@@

\subsection*{flowCore basic data structures}
\subsubsection*{flowFrame}
\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other successful Bioconductor
packages. All information from a single FCS file, i.e., the collection
of events and the accompanying metadata, is stored in one single
container. We call the structure that hold this data a
\Rclass{flowFrame}. Raw data values as well as associated metadata of
a \Rclass{flowFrame} can be accessed programmatically. Most commonly,
the metadata consist of descriptors of the stains used in the
experiment and the respective measurement channels, information about
compensation performed at the instrument side and any additional
keywords the user deems to be important to annotate the data. During
the creation of a \Rclass{flowFrame}, a number of quality checks are
performed to ensure data integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple samples, hence we introduce the
concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
\Rclass{flowFrames} along with relevant information associated with
each individual frame such as descriptions of the cell sample, the
treatment to which the sample was subjected, or the location of that
sample in a 96 or 384 well plate.  \Rclass{flowSets} manage the
consistent application of operations on the individual
\Rclass{flowFrames} and shift the burden of keeping score of the
metadata from the user to the infrastructure, thus reducing the risk
of errors (\textit{e.g.}, mixups of sample labels). The
\Rclass{flowSet} structure can be readily extended to incorporate the
potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials, where hundreds of patients might
provide samples at different time points over the course of the
experiment.

\subsection*{Standard flow operations}

Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, typically in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of
\Rpackage{flowCore} through object oriented programming are features
that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

\subsubsection*{Transformation and compensation}

Data transformation is essential for both data visualization and
modelling \cite{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantial or arcsinh,
see Table~\ref{table1} for a complete list). Furthermore, the design
of the R language makes it easy to define arbitrary functions to apply
to the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation is available for both
\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate single stain controls.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
  \subsection*{From quality assessment to batch gating}
  The \Rpackage{flowCore} package has been successfully applied in the
  analysis of several datasets
  \cite{gasparetto2004ice,brinkman2007hcf}. A complete documentation
  of the \Rpackage{flowCore} software is beyond the scope of this
  publication.  Much more comprehensive documentation and users guide
  information with programmatic examples is available online
  (\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
  and as part of the package distribution. Here, we want to briefly
  exemplify some of the software's key features, that is, the coherent
  treatment of all samples in a potentially large experiment, the
  concept of data-driven automated gating, the integration of existing
  software into the framework, and the generation of
  publication-quality graphics for data visualization.

@@

\subsection*{Related flow packages}

In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \cite{sarkar2008ufv}
provides sophisticated data visualization tools, that make use off
multivariate trellis plotting \cite{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results. The \Rpackage{flowUtil} package implements data import and
export including flow-cytometry specific standard markup
language. Finally, the \Rpackage{flowStats} package provides elaborate
statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust}. Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitating the
handling of spatial metadata, is under development.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
Through \Rpackage{flowCore}, we have provided the FCM community with
an open source, freely available, highly functional, standards
compliant development and analysis platform for high throughput data
analysis.  Our intention is to provide a platform for the
collaborative development of new analysis methods that will facilitate
the transition of these new methods to the larger flow community.  Our
experience has been that the collaborative effort we enable for
devising new methodology has been proven beneficial for a number of
different biological and computational biology challenges.  We hope
that our framework will be the foundation for fruitful shared research
by many collaborators from multiple scientific fields and will help
resolve bottlenecks that currently prevent further development and
deployment of FC-HCS to increasingly complex and important scientific
and clinical applications.

>>>>

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
  \paragraph*{Background:} Recent advances in automation technologies
  have enabled the use of flow-cytometry high content screening
  (FH-HCS), generating large complex data sets often in clinical
  trials or drug discovery settings. However, data management and data
  analysis methods have not advanced sufficiently far from the initial
  small-scale studies to support modeling in the presence of multiple
  covariates.

  \paragraph*{Results:} We developed a set of flexible open source
  computational tools (in the R package \Rpackage{flowCore}) to
  facilitate the analysis of these complex data. A key component of
  which is having suitable data structures that support the
  application of similar operations to a collection of samples or a
  clinical cohort. In addition, our software constitutes a shared and
  extensible research platform that enables colaboration between
  bioinformaticians, computer scientists, statisticians, biologists
  and clinicians. This platform will foster the developement of novel
  analytic methods for flow cytometry.

  \paragraph*{Conclusions:} 
  The software has been applied in the analysis of various data sets
  and its data structures have been proven to be highly efficient in
  capturing and organizing the analytic work flow. A number of
  additional Bioconductor packages successfully build on the
  infrastructure provided by \Rpackage{flowCore}, opening new avenues
  for flow data analysis.

\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Automation technologies developed during the last several years have
enabled the use of flow cytometry high content screening (FC-HCS) to
generate large, complex data sets in both basic and clinical research
applications \cite{brinkman2007hcf}. A serious bottleneck in the
interpretation of existing studies and the application of FC-HCS to
even larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of flow cytometry (FCM) to small-scale,
tube-based studies \cite{mahnke2007omi}. In particular, the data often
need to be organized into groups of samples based on combinations of
additional covariates and similar operations need to be applied to
these groups in a transparent and reproducible manner. Furthermore,
the growing depth of knowledge in the field of immunology, for
instance the charactarization of distinct human T-cell subpopulation
\cite{Appay}, clearly argues for more systematic approaches.
%% add one of the last roederer's paper that "request" algo and
%% automation for FC-HCS data analysis.

Some of the consequences of the lag of efficient software solutions
are difficulties in maintaining the integrity and documentation of
large data sets, assessing measurement quality, developing validated
assays, controlling the accuracy of gating techniques, automating
complex gating strategies, and aggregating statistical results across
large study sets for further analysis. In addition, new analysis
approaches face difficulty in finding their way into standard
practice. We believe that these barriers to the development and
dissemination of new analysis methods is one of the fundamental
restraints on the future expansion of FC-HCS in both clinical and
research applications.

Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. Most of the currently available
analysis tools are designed to reflect this work flow.  We believe
that these approaches, in addition to being expensive and labor
intensive, do not fully address the highly complex nature of FCM data;
in particular, they disregard many of the fundamental aspects of the
data, such as sample groups or cohorts, the underlying distribution or
its high-dimensional nature. Furthermore, the subjective character of
manual analyses are a major obstacle to reproducibility. In a recent
study of flow cytometric standardization involving 15 institutions,
the mean inter-laboratory coefficient of variation ranged from 17 to
44\%, even though preparation was standardized and performed using the
same samples and reagents at each site \cite{Maecker2005}. For FC-HCS
data, unassisted manual inspection is extremely time consuming, and
robust statistical methods need to be developed to point investigators
to interesting aspects of the data, or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that collaboration with
other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

In this paper, we describe a set of flexible and well structured
computational tools to efficiently analyze FC-HCS data. Our intent is
to provide a shared research platform that enables bioinformaticians,
computer scientists, and statisticians to work collaboratively with
biologists and clinicians to develop novel methods for FCM data
analysis, a process deemed crucial by many for the further development
of the technology \cite{lizard2007fca}.

@@

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
The computational tools we have developed are distributed in the R
software language \cite{Rmain} as the Bioconductor \cite{BIOC} package
\Rpackage{flowCore}.  The package \Rpackage{flowCore} is a freely
available, highly functional, and extensible FCM data analysis
platform that enables researchers to efficiently handle FC-HCS data
and encourages open development of tools for their coherent
analysis. In our implementation of \Rpackage{flowCore} we rely on two
important lessons learned from the field of gene expression data
analysis: the first being the importance of data structures that
reflect the underlying data and facilitate the manipulations that are
of most interest, while the second is the importance of a modular
architecture that allows for many developers to extend and use the
underlying infrastructure and to combine tools in complex work flows.
\Rpackage{flowCore} implements such computationally efficient data
structures and a range of specialized methods addressing all
components of a typical FCM analysis work flow, including
compensation, transformation, and gating. \Rpackage{flowCore} runs on
Windows, Mac OS X, and Linux/Unix operating systems. 
%% In this paper, we describe the functionality and methods underlying
%% \Rpackage{flowCore}, and provide examples of their use.

@@

\subsection*{Basic Data Structures}
\subsubsection*{flowFrame}
\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other successful Bioconductor
packages. All information from a single FCS file, i.e., the collection
of events and the accompanying metadata, is stored in one single
container. We call the structure that hold this data a
\Rclass{flowFrame}. Raw data values as well as associated metadata of
a \Rclass{flowFrame} can be accessed programmatically. Most commonly,
the metadata consist of descriptors of the stains used in the
experiment and the respective measurement channels, information about
compensation performed at the instrument side and any additional
keywords the user deems to be important to annotate the data. During
the creation of a \Rclass{flowFrame}, a number of quality checks are
performed to ensure data integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple samples, hence we introduce the
concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
\Rclass{flowFrames} along with relevant information associated with
each individual frame such as descriptions of the cell sample, the
treatment to which the sample was subjected, or the location of that
sample in a microtitre plate.  \Rclass{flowSets} manage the consistent
application of operations on the individual \Rclass{flowFrames} and
shift the burden of keeping score of the metadata from the user to the
infrastructure, thus reducing the risk of errors (\textit{e.g.},
mixups of sample labels). Crucial operations like taking subsets, data
transformations and gating or computation of summary statistics are
greatly facilitated and all relevant annotation information is
constantly passed on along the analysis pipeline.

The \Rclass{flowSet} structure can be readily extended to incorporate
the potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials, where hundreds of patients might
provide samples at different time points over the course of the
experiment. The \Rclass{flowSet} data structure is one of the key
features in the \Rpackage{flowCore} package and it is fundamental to
the implementation of many of the high level functionalities such as
quality assessment and control, visualization and automateds gating.

\subsection*{Standard flow operations}
Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, typically in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of
\Rpackage{flowCore} through object oriented programming are features
that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

\subsubsection*{Transformation and compensation}
Data transformation is essential for both data visualization and
modelling \cite{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantial or arcsinh,
see Table~\ref{table1} for a complete list). Furthermore, the design
of the R language makes it easy to define arbitrary functions to apply
to the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation is available for both
\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate single stain controls.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
\subsection*{From quality assessment to batch gating}
The \Rpackage{flowCore} package has been successfully applied in the
analysis of several data sets, both originating from clinical trials
\cite{brinkman2007hcf} and drug discovery experiments
\cite{gasparetto2004ice}. A complete description of the
\Rpackage{flowCore} software is beyond the scope of this publication.
Much more comprehensive documentation and users guide information with
programmatic examples is available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
and as part of the package distribution. Here, we want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large experiment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

@@

\subsection*{Related flow packages}
In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \cite{sarkar2008ufv}
provides sophisticated data visualization tools, that make use off
multivariate trellis plotting \cite{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results. The \Rpackage{flowUtil} package implements data import and
export including flow-cytometry specific standard markup
language. Finally, the \Rpackage{flowStats} package provides elaborate
statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust}. Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitating the
handling of spatial metadata, is under development.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
Through \Rpackage{flowCore}, we have provided the FCM community with
an open source, freely available, highly functional, standards
compliant development and analysis platform for high throughput data
analysis.  We hope to foster collaborative development of new analysis
methods and to facilitate the transition of these new methods to the
larger flow community.  Our experience has been that such
collaborative effort has proven beneficial for a number of different
biological and computational biology challenges, greatly elevating
their applicability.  We hope that our framework will be the
foundation for fruitful shared research by many collaborators from
multiple scientific fields and will help resolve bottlenecks that
currently prevent further development and deployment of FC-HCS to
increasingly complex and important scientific and clinical
applications.
